---
title: "FEDUP manual"
author: "Catherine Ross"
package: FEDUP
vignette: >
  %\VignetteIndexEntry{FEDUP manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`FEDUP` is an R package for gene enrichment and depletion analysis on
user-defined pathways using a Fisher's exact test.

## Getting started
### System prerequisites

R version > 3.5.0

R packages:

* **CRAN**: openxlsx, tibble, dplyr, data.table, ggplot2, ggthemes, forcats,
            RColorBrewer  
* **Bioconductor**: RCy3

### Installation

Install FEDUP via devtools

```{r, install_FEDUP, message = FALSE, warning = FALSE}
#devtools::install_github("rosscm/FEDUP")
devtools::load_all()
```

## Quick run
### Data input

Load example test genes, background genes, and pathways

To note, the test genes comprise solely of the **olfactory signaling** pathway. As such, we would expect to see strong *enrichment* for olfactory-associated pathways and *depletion* for pathways not associated with olfactory signaling.

```{r, load_input}
data(testGene)
data(backgroundGene)
data(pathwaysGMT)
```

Take a look at the input data structure

```{r, input_structure}
str(testGene)
str(backgroundGene)
str(pathwaysGMT)
```

Run FEDUP on sample data

```{r, run_FEDUP}
fedup_res <- runFedup(testGene, backgroundGene, pathwaysGMT)
```

View of output results table sorted by p-value

```{r, view_FEDUP}
print(fedup_res)
```

### Visualization

Plot top pathway enrichment/depletion results

```{r, plot_FEDUP, fig.width = 10, fig.height = 7}
fedup_plot <- fedup_res[which(fedup_res$fdr < 0.05),]
fedup_plot$log10fdr <- -log10(fedup_plot$fdr + 1e-10) # log10-transform FDR for plotting
fedup_plot$pathway <- gsub("\\%.*", "", fedup_plot$pathway) # clean pathway names
plotDotPlot(
   df = fedup_plot,
   x_var = "log10fdr",
   y_var = "pathway",
   x_lab = "-log10(FDR)",
   fill_var = "enrichment",
   fill_lab = "Enrichment",
   size_var = "real_pathway_frac",
   size_lab = "Gene fraction"
)
```

## Versioning

For the versions available, see the [tags on this repo](https://github.com/rosscm/FEDUP/tags).
